---
title: "Describing Numerical and Categorical Data"
subtitle: "<br><br> STA 032: Gateway to data science Lecture 10"
author: "Jingwei Xiong"
date: April 24, 2023
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts]
    # css: ["default", "extra.css"]
    # css: ["../xaringan-themer.css", "../slides.css"]
    # lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
---

```{r, echo = FALSE, eval = FALSE}
library(renderthis)
to_pdf(from = "lecture10.html",complex_slides = TRUE)
```


```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons"))
xaringanExtra::use_panelset()
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = "",eval = TRUE,fig.retina = 2, message=F, warning = F, fig.height = 5.5
)
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(unvotes)
library(tidyverse)
library(openintro)
```


```{css, echo = FALSE}
.tiny .remark-code { font-size: 60%; }
.small .remark-code { font-size: 80%; }
```



```{r include = FALSE}
loans <- loans_full_schema %>%
  select(loan_amount, interest_rate, term, grade, 
         state, annual_income, homeownership, debt_to_income,
         issue_month)
```

## Reminders

- HW 2 due April 26 12pm.

- HW 3 due May 1 12pm.
  - Please start the homework as soon as possible.
  
- **Midterm 1** (Open book, take home, approximate 2 hours, time limit 5 hours) 
  - **Due April 29 midnight, cover lecture 1-12**
  - **Will not cover the Level 2 part of Homework 3**
  - Preparing guides: Finish Homework 1-3, be familiar to the lecture slides.
  - You can copy any your own codes in homework 1-3 to finish the open book exam.
  
- The PDF version of slides can be found on Canvas. 

  - You can use search on it to find function examples.

---

