---
title: "Describing numerical distributions"
subtitle: "<br><br> STA 032: Gateway to data science Lecture 9"
author: "Jingwei Xiong"
date: April 21, 2023
output:
  xaringan::moon_reader:
    # css: [default, metropolis, metropolis-fonts]
    # css: ["default", "extra.css"]
    # css: ["../xaringan-themer.css", "../slides.css"]
    # lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
---

```{r, echo = FALSE, eval = FALSE}
library(renderthis)
to_pdf(from = "lecture9.html",complex_slides = TRUE)
```


```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons"))
xaringanExtra::use_panelset()
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = "",eval = TRUE,fig.retina = 2, message=F, warning = F
)
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(unvotes)
```


```{css, echo = FALSE}
.tiny .remark-code { font-size: 60%; }
.small .remark-code { font-size: 80%; }
```


## Reminders

- HW 2 due April 26 12pm.

- HW 3 due May 1 12pm.
  - Please start the homework as soon as possible.
  
- **Midterm 1** (Open book, take home, approximate 2 hours, time limit 5 hours) 
  - **Due April 29 midnight, cover lecture 1-12**
  - Preparing guides: Finish Homework 1-3, be familiar to the lecture slides.
  - You can copy any your own codes in homework 1-3 to finish the open book exam.

---

## Today

- Facets

- Time series

- Descriptive statistics


---

